---
title: "Dataset comparisons"
author: "Dan Bunis"
date: "9/28/2020"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dittoSeq)
library(DESeq2)
library(Seurat)
library(MAST)
library(eulerr)
```

# NOTE: This code requires outputs of other markdown code from this porject and should be run last.

Required data:

- Final output, bulkCD4.Rds, of the bulk naive CD4 T cell analysis (bulkRNAseq_CD4naiveTcells.Rmd).
- Final output, Tcells.Rds, of the single-cell naive T cell analyses (scRNAseq_naiveTcells.Rmd)
- Differential expression output of T cells, Microarray_APBvsFPBTcells_FDR0.05_FC1.5_LongAnn.csv, from microarray analysis (generated in the first half of microarray-and-qRTPCR_CD4naiveTcells-and-monocytes.Rmd)

# 0) Load datasets
```{r}
Tcells <- readRDS("Tcells.rds")
bulkCD4 <- readRDS("bulkCD4.rds")
bulkCD4_dds <- readRDS("bulkCD4_dds.rds")
```

# 1) Comparison of bulk and single-cell naive CD4 T cell differential expression calls

Equivalent differential expression cutoff:

- Absolute Fold Change >= 1.5
- FDR <= 0.05
- Non-ribosomal & non-mitochondrial (not analyzed in the bulk RNAseq datasset)

```{r, results = "hide"}
# Create function to remove ribosomal and mitochondrial genes
remove.ribo.mito <- function(genelist){
  if (length(grep("^RPL|^RPS|^MT-",genelist))>0){
    return(genelist[-grep("^RPL|^RPS|^MT-",genelist)])
  } else {
    genelist
  }
}

# Set the clustering to be by T-celltype + age ("Tage" meta) so the Seurat FindMarkers function will use this.
Idents(Tcells) <- "Tage"
# Fetal vs Adult
sc4FvA <- FindMarkers(Tcells,
                     ident.1 = "4-fetal",
                     ident.2 = "4-adult",
                     test.use = "MAST")
sc4FvA_padjFC <- sc4FvA[abs(sc4FvA$avg_logFC)>=0.585 &
                          sc4FvA$p_val_adj<=0.05 &
                          !is.na(sc4FvA$p_val_adj),]
# Fetal vs UCB
sc4FvC <- FindMarkers(Tcells,
                     ident.1 = "4-fetal",
                     ident.2 = "4-cord",
                     test.use = "MAST")
sc4FvC_padjFC <- sc4FvC[abs(sc4FvC$avg_logFC)>=0.585 &
                          sc4FvC$p_val_adj<=0.05 &
                          !is.na(sc4FvC$p_val_adj),]
# UCB vs Adult
sc4CvA <- FindMarkers(Tcells,
                     ident.1 = "4-cord",
                     ident.2 = "4-adult",
                     test.use = "MAST")
sc4CvA_padjFC <- sc4CvA[abs(sc4CvA$avg_logFC)>=0.585 &
                          sc4CvA$p_val_adj<=0.05 &
                          !is.na(sc4CvA$p_val_adj),]

### Split up and down genes.
FETAL_sc4FvA <- remove.ribo.mito(rownames(sc4FvA_padjFC[sc4FvA_padjFC$avg_logFC>0,]))
ADULT_sc4FvA <- remove.ribo.mito(rownames(sc4FvA_padjFC[sc4FvA_padjFC$avg_logFC<0,]))

FETAL_sc4FvC <- remove.ribo.mito(rownames(sc4FvC_padjFC[sc4FvC_padjFC$avg_logFC>0,]))
UCB_sc4FvC <- remove.ribo.mito(rownames(sc4FvC_padjFC[sc4FvC_padjFC$avg_logFC<0,]))

UCB_sc4CvA <- remove.ribo.mito(rownames(sc4CvA_padjFC[sc4CvA_padjFC$avg_logFC>0,]))
ADULT_sc4CvA <- remove.ribo.mito(rownames(sc4CvA_padjFC[sc4CvA_padjFC$avg_logFC<0,]))
```

Re-calculate the bulk markers

```{r, results = "hide"}
#Extract comparisons from DESeq
FvA <- as.data.frame(results(bulkCD4_dds, contrast=c("Age","Fetal","Adult")))
FvC <- as.data.frame(results(bulkCD4_dds, contrast=c("Age","Fetal","Cord")))
CvA <- as.data.frame(results(bulkCD4_dds, contrast=c("Age","Cord","Adult")))

#Trim to DE cutoffs
FvA_padjFC <- FvA[(
   (FvA$padj<=0.05) & (abs(FvA$log2FoldChange)>=0.585*log2(exp(1))) & (!is.na(FvA$padj))
   ),]
FvC_padjFC <- FvC[(
  (FvC$padj<=0.05) & (abs(FvC$log2FoldChange)>=0.585*log2(exp(1))) & (!is.na(FvC$padj))
  ),]
CvA_padjFC <- CvA[(
  (CvA$padj<=0.05) & (abs(CvA$log2FoldChange)>=0.585*log2(exp(1))) & (!is.na(CvA$padj))
  ),]

### Split up and down genes.
FETAL_FvA <- rownames(FvA_padjFC[FvA_padjFC$log2FoldChange>0,])
FETAL_FvC <- rownames(FvC_padjFC[FvC_padjFC$log2FoldChange>0,])

UCB_FvC <- rownames(FvC_padjFC[FvC_padjFC$log2FoldChange<0,])
UCB_CvA <- rownames(CvA_padjFC[CvA_padjFC$log2FoldChange>0,])

ADULT_FvA <- rownames(FvA_padjFC[FvA_padjFC$log2FoldChange<0,])
ADULT_CvA <- rownames(CvA_padjFC[CvA_padjFC$log2FoldChange<0,])
```

## Venns

```{r}
width = 2.5
height = 2.3

eulerr_options(fills = list(fill = dittoColors()[c(4,5)],
                            alpha = c(1,1)),
               labels = list(fontsize = 9))
set.seed(1908)
pdf("Comparison-Figs/venn_CD4s_FetalvA.pdf", w=width, h=height)
plot(venn(c('bulkCD4s' = list(FETAL_FvA),
            'scCD4s' = list(FETAL_sc4FvA))),
     main = "Fetal Enriched")
dev.off()
set.seed(1908)
pdf("Comparison-Figs/venn_CD4s_AdultvF.pdf", w=width, h=height)
plot(venn(c('bulkCD4s' = list(ADULT_FvA),
            'scCD4s' = list(ADULT_sc4FvA))),
     main = "Adult Enriched")
dev.off()
set.seed(1908)
pdf("Comparison-Figs/venn_CD4s_FetalvC.pdf", w=width, h=height)
plot(venn(c('bulkCD4s' = list(FETAL_FvC),
            'scCD4s' = list(FETAL_sc4FvC))),
     main = "Fetal Enriched")
dev.off()
set.seed(1908)
pdf("Comparison-Figs/venn_CD4s_CordvF.pdf", w=width, h=height)
plot(venn(c('bulkCD4s' = list(UCB_FvC),
            'scCD4s' = list(UCB_sc4FvC))),
     main = "UCB Enriched")
dev.off()
set.seed(1908)
pdf("Comparison-Figs/venn_CD4s_CordvA.pdf", w=width, h=height)
plot(venn(c('bulkCD4s' = list(UCB_CvA),
            'scCD4s' = list(UCB_sc4CvA))),
     main = "UCB Enriched")
dev.off()
set.seed(1908)
pdf("Comparison-Figs/venn_CD4s_AdultvC.pdf", w=width, h=height)
plot(venn(c('bulkCD4s' = list(ADULT_CvA),
            'scCD4s' = list(ADULT_sc4CvA))),
     main = "Adult Enriched")
dev.off()
```

## Overlapping Genes Exploration

```{r}
allsc4 <- unique(c(rownames(sc4FvA_padjFC),
                   rownames(sc4FvC_padjFC),
                   rownames(sc4CvA_padjFC)))
allbulk4 <- unique(c(FETAL_FvA, FETAL_FvC,
                     UCB_FvC, UCB_CvA,
                     ADULT_FvA, ADULT_CvA))

# What are the unique-to-single-cell genes in the Fetal vs Adult comparison?
remove.ribo.mito(rownames(
  sc4FvA_padjFC[!rownames(sc4FvA_padjFC) %in% c(FETAL_FvA, ADULT_FvA),]))
# What are the unique-to-single-cell genes in the Fetal vs UCB comparison?
remove.ribo.mito(rownames(
  sc4FvC_padjFC[!rownames(sc4FvC_padjFC) %in% c(FETAL_FvC, UCB_FvC),]))
```

# 2) Compare expression of peripheral-only Microarray FvA genes in mixed tissue bulk RNAseq

This is a comparison of how the naive CD4 T cell gene signature obtained from comparison of fetal peripheral versus adult peripheral blood samples with microarray works for characterizing Fetal-splenic (vs UCB) vs Adult peripheral blood naive CD4 T cells in bulk RNA-seq.

```{r}
# Read in microarray signature
# The differential expression microaarray data is already cutoff to log2FC=1.5 and FDR <= 0.05.
microarray <- read.csv("Microarray_APBvsFPBTcells_FDR0.05_FC1.5_LongAnn.csv",
                       header = T, row.names = 1)
# Remove genes not analyzed further in this dataset.
microarray <- microarray[-grep("^XLOC|^LOC|^ENST|ORF|orf|A_19|A_24|A_33", rownames(microarray)),]

# Trim to only the genes in the bulk RNAseq dataset.
microarray_in <- microarray[isGene(rownames(microarray),bulkCD4),]

# Store directionality
enriched <- array("fetal-enriched", nrow(microarray_in))
enriched[microarray_in$logFC>0] <- "adult-enriched"

# Create a metadata for easily labeling the samples with their tissue identities.
age.names <- factor(meta("Samples", bulkCD4), labels = paste0(rep(c("APB-","FS-","UCB-"),each = 5), 1:5))
bulkCD4$age.names <- age.names

### Plot
# Create Heatmap
hm <- dittoHeatmap(
  genes = rownames(microarray_in), object = bulkCD4, annot.by = "Age",
  annotation_row = data.frame(
    'MicroarrayAnnotation' = enriched,
    row.names = rownames(microarray_in)),
  annot.colors = c(dittoColors()[1:3], "red", "blue"),
  cutree_row = 2, cutree_col = 3,
  show_rownames = FALSE, cell.names.meta = "age.names",
  treeheight_row = 15, treeheight_col = 15)
# Create fetal (microarray) gene summary plot
(pF <- dittoPlotVarsAcrossGroups(
  rownames(microarray_in[microarray_in$logFC<0,]),
  object = bulkCD4,
  group.by = "Age",
  x.reorder = 3:1, colors = 3:1,
  x.labels = c("FS", "UCB", "APB"),
  plots = c("vlnplot","boxplot"),
  adjustment = "z-score",
  main = NULL, jitter.size = 0.3,
  sub = "Fetal-enriched\nmicroarray genes",
  boxplot.color = "white",
  boxplot.fill = FALSE,
  y.breaks = -1:1,
  xlab = NULL,
  ylab = NULL,
  legend.show = FALSE,
  vlnplot.lineweight = 0.5,
  theme = theme_bw()))
# Create adult (microarray) gene summary plot
(pA <- dittoPlotVarsAcrossGroups(
  rownames(microarray_in[microarray_in$logFC>0,]),
  object = bulkCD4,
  group.by = "Age",
  x.reorder = 3:1, colors = 3:1,
  x.labels = c("FS", "UCB", "APB"),
  plots = c("vlnplot","boxplot"),
  adjustment = "z-score",
  main = NULL, jitter.size = 0.3,
  sub = "Adult-enriched\nmicroarray genes",
  boxplot.color = "white",
  boxplot.fill = FALSE,
  y.breaks = -1:1,
  xlab = NULL,
  ylab = NULL,
  legend.show = FALSE,
  vlnplot.lineweight = 0.5,
  theme = theme_bw()))
# Plot these altogether.
pdf("Comparison-Figs/RNAseqVsMicroarray_heatmap.pdf", w = 7, h = 6)
gridExtra::grid.arrange(grobs = list(hm[[4]],pF,pA),
    ncol = 7,
    nrow = 5,
    layout_matrix = 
        rbind(c(1,1,1,1,1,2,2),
              c(1,1,1,1,1,2,2),
              c(1,1,1,1,1,3,3),
              c(1,1,1,1,1,3,3),
              c(1,1,1,1,1,NA,NA)))
dev.off()
```

# 3) Compare expression in bulk RNAseq samples to Treg signature from Ng, et al. 2019.

For this comparison, we need some supplemental data from Ng et. al., "Helios enhances the preferential differentiation of human fetal CD4+ naïve T cells into regulatory T cells", Science, 2019:

- Table S1: https://immunology.sciencemag.org/highwire/filestream/642275/field_highwire_adjunct_files/1/aav5947_Table_S1.xlsx

The table contains two tabs, but we need data from both, so I copied the second tab into the first, remove the extra cell of data holding legend information, then exported as a csv.  That csv is what gets loaded in in the first line of code below.

```{r}
Treg_sig <- read.csv("Ng2019_Table_S1_TregSignature_bothTabs.csv",
                     stringsAsFactors = F, header = T)[,c(1,8)]
# Extract Treg up genes captured in this dataset
Treg_UP <- Treg_sig$Gene_symbol[Treg_sig$Heatmap.Cluster %in% c("Cluster 1","Cluster 2")]
Treg_UP_in <- Treg_UP[isGene(Treg_UP, bulkCD4)]
# Extract Treg down genes captured in this dataset
Treg_DOWN <- Treg_sig$Gene_symbol[Treg_sig$Heatmap.Cluster %in% c("Cluster 3","Cluster 4")]
Treg_DOWN_in <- Treg_DOWN[isGene(Treg_DOWN, bulkCD4)] 

#Make age annotation metadata
bulkCD4$Age.names <- paste(gsub("Cord", "UCB",meta("Age", bulkCD4)),1:5, sep = "-")

#Make Treg gene annotations:
genes_annot <- data.frame(TregExpression = c(rep("Up", length(Treg_UP_in)),
                                             rep("Down", length(Treg_DOWN_in))),
                          row.names = c(Treg_UP_in, Treg_DOWN_in),
                          stringsAsFactors = FALSE)

# Select specific Treg genes to highlight:
highlights <- c("IL2RA", "FOXP3", "CTLA4", "IKZF2", "IKZF4")
# highlights_in <- isGene(highlights, bulkCD4, TRUE) # ALl are inside
# Collect average z-score expression levels of each gene in all ages, then separate by gene for adding these to the summary plots
all_exp_df <- data.frame(
  z.exp = 
    c(sapply(highlights, function(X) {
      mean(gene(X, bulkCD4, adjustment = "z-score")[meta("Age",bulkCD4)=="Fetal"])
    }),
    sapply(highlights, function(X) {
      mean(gene(X, bulkCD4, adjustment = "z-score")[meta("Age",bulkCD4)=="Cord"])
    }),
    sapply(highlights, function(X) {
      mean(gene(X, bulkCD4, adjustment = "z-score")[meta("Age",bulkCD4)=="Adult"])
    })),
  gene = rep(highlights,3),
  grouping = rep(c("F","UCB","A"), each = 5))
cord_exp_df <- all_exp_df[all_exp_df$grouping == "UCB",]

# Make plots
hm <- dittoHeatmap(rownames(genes_annot),
    object = bulkCD4,
    annot.by = "Age",
    cell.names.meta = "Age.names",
    main = "Treg Signature Genes",
    annotation_row = genes_annot,
    cutree_row = 2, treeheight_row = 15,
    cutree_col = 3, treeheight_col = 15,
    highlight.genes = highlights,
    show_rownames = FALSE,
    annot.colors = c(dittoColors()[1:3],"blue","red"))
(pUP <- dittoPlotVarsAcrossGroups(
    c(Treg_UP_in),
    object = bulkCD4,
    group.by = "Age",
    x.reorder = 3:1, colors = 3:1,
    x.labels = c("F", "UCB", "A"),
    plots = c("vlnplot","boxplot"),
    adjustment = "z-score",
    main = NULL, jitter.size = 0.3,
    sub = "Treg-upregulated",
    boxplot.color = "white",
    boxplot.fill = FALSE,
    y.breaks = -1:1,
    theme = theme_bw(),
    x.labels.rotate = FALSE,
    ylab = NULL, xlab = NULL,
    legend.show = FALSE,
    vlnplot.lineweight = 0.5) +
    # Add gene annotations
    geom_point(
        data = cord_exp_df,
        aes(x = grouping, y = z.exp),
        fill = "black",
        shape = 15,
        inherit.aes = FALSE) +
    ggrepel::geom_text_repel(
        data = cord_exp_df,
        aes(x = grouping, y = z.exp, label = gene),
        fill = "white",
        direction = "x",
        inherit.aes = FALSE)# +
    # geom_path(
    #     data = all_exp_df,
    #     aes(x = grouping, y = z.exp, group = gene),
    #     inherit.aes = FALSE)
    )
(pDN <- dittoPlotVarsAcrossGroups(
    c(Treg_DOWN_in),
    object = bulkCD4,
    group.by = "Age",
    x.reorder = 3:1, colors = 3:1,
    x.labels = c("F", "UCB", "A"),
    plots = c("vlnplot","boxplot"),
    adjustment = "z-score",
    main = NULL, jitter.size = 0.3,
    sub = "Treg-downregulated",
    boxplot.color = "white",
    boxplot.fill = FALSE,
    y.breaks = -1:1,
    theme = theme_bw(),
    x.labels.rotate = FALSE,
    ylab = NULL, xlab = NULL,
    legend.show = FALSE,
    vlnplot.lineweight = 0.5))
pdf("Comparison-Figs/bulkCD4s_TregSignature_heatmap.pdf", w = 8, h = 4.5)
gridExtra::grid.arrange(grobs = list(hm[[4]],pUP,pDN),
    ncol = 7,
    nrow = 4,
    layout_matrix = 
        rbind(c(1,1,1,1,1,1,2,2),
              c(1,1,1,1,1,1,2,2),
              c(1,1,1,1,1,1,3,3),
              c(1,1,1,1,1,1,3,3)))
dev.off()
```

```{r}
# Statistics
df_TregUP <- dittoPlotVarsAcrossGroups(
    c(Treg_UP_in),
    object = bulkCD4,
    group.by = "Age",
    adjustment = "z-score",
    data.out = TRUE)[[2]]
df_TregDOWN <- dittoPlotVarsAcrossGroups(
    c(Treg_DOWN_in),
    object = bulkCD4,
    group.by = "Age",
    adjustment = "z-score",
    data.out = TRUE)[[2]]
# Up
wilcox.test(df_TregUP$var.data[df_TregUP$grouping=="Cord"], df_TregUP$var.data[df_TregUP$grouping=="Adult"])
wilcox.test(df_TregUP$var.data[df_TregUP$grouping=="Cord"], df_TregUP$var.data[df_TregUP$grouping=="Fetal"])
# Down
wilcox.test(df_TregDOWN$var.data[df_TregDOWN$grouping=="Cord"], df_TregDOWN$var.data[df_TregDOWN$grouping=="Adult"])
wilcox.test(df_TregDOWN$var.data[df_TregDOWN$grouping=="Cord"], df_TregDOWN$var.data[df_TregDOWN$grouping=="Fetal"])
```


