---
title: "Naive CD4 T cell & classical monocyte microarray and qRT-PCR analysis"
author: "Dan Bunis"
date: "2/26/2020"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(reshape2)
library(ggplot2)
library(ggrepel)
library(limma)
library(dittoSeq) # For its colors
library(GEOquery)
```

First, some functions
```{r}
mydistfunPearson = function(x) {
t.dist = 1 - cor(t(x), use="pairwise")
t.limit <- 1.1*max(t.dist,na.rm=TRUE)
t.dist[is.na(t.dist)] = t.limit
t.dist = as.dist(t.dist)
return(t.dist)
}

mydistfunEuc = function(x) {
t.dist = dist(x)
t.limit <- 1.1*max(t.dist,na.rm=TRUE)
t.dist[is.na(t.dist)] = t.limit
return(t.dist)
}

outliersNA <- function(x,p) {
  z <- (x-mean(x,na.rm=TRUE)) / sd(x,na.rm=TRUE)
  rm.index <- which( pnorm(abs(z), lower.tail=FALSE) < p)
  x[rm.index] <- NA
  return(x)
}

summarySE <- function(data=NULL, measurevar, groupvars=NULL, na.rm=FALSE,
                      conf.interval=.95, .drop=TRUE) {
  ##############################################
  ### Function obtained on July 20, 2019, from:
  # http://www.cookbook-r.com/Graphs/Plotting_means_and_error_bars_(ggplot2)/
  ##############################################
  
  ## Gives count, mean, standard deviation, standard error of the mean, and
  ## confidence interval (default 95%).
  ##   data: a data frame.
  ##   measurevar: the name of a column that contains the variable to be summariezed
  ##   groupvars: a vector containing names of columns that contain grouping variables
  ##   na.rm: a boolean that indicates whether to ignore NA's
  ##   conf.interval: the percent range of the confidence interval (default is 95%)
  library(plyr)
  
  # New version of length which can handle NA's: if na.rm==T, don't count them
  length2 <- function (x, na.rm=FALSE) {
    if (na.rm) sum(!is.na(x))
    else       length(x)
  }
  
  # This does the summary. For each group's data frame, return a vector with
  # N, mean, and sd
  datac <- ddply(data, groupvars, .drop=.drop,
                 .fun = function(xx, col) {
                   c(N    = length2(xx[[col]], na.rm=na.rm),
                     mean = mean   (xx[[col]], na.rm=na.rm),
                     sd   = sd     (xx[[col]], na.rm=na.rm)
                   )
                 },
                 measurevar
  )
  
  # Rename the "mean" column    
  datac <- rename(datac, c("mean" = measurevar))
  
  datac$se <- datac$sd / sqrt(datac$N)  # Calculate standard error of the mean
  
  # Confidence interval multiplier for standard error
  # Calculate t-statistic for confidence interval: 
  # e.g., if conf.interval is .95, use .975 (above/below), and use df=N-1
  ciMult <- qt(conf.interval/2 + .5, datac$N-1)
  datac$ci <- datac$se * ciMult
  
  return(datac)
}
```

# 1. Check for outliers & run differential expression on microarrays

## T cells first

### 0) Load in the data and get started

```{r}
# Load most raw form and median scale
cluster1=read.csv("Microarray_annotatedData/17_medGen_longAnn_Tcells.csv", header=T,row.names = 1,check.names=FALSE)
```

### 1) Check for outliers:

```{r}
#### CLUSTER FPB AND APB - T cells
cluster2=cluster1[,c(4:13)]
distance = dist(t(cluster2), method="euclidean")
hc <- hclust(distance, method="average")
plot(hc, main = "Cluster Dendrogram", ylab="Height", xlab =NULL)
```

Fetal peripheral blood #5 clusters on its own here and in the monocytes analysis.

Differential Expression was run then for T cells (and monocytes) without FPB#5.

### 2) Select best probes (highest interquartile range) for genes with multiple.

```{r}
cluster2=cluster1[,c(1,2,4:13)]
cluster2$NAME = sapply(as.character(cluster2$NAME),function(x) strsplit(x," ")[[1]][1])

# pick the most variable probe for each replicated gene
replicated = unique(cluster2$NAME[duplicated(cluster2$NAME)])
# separate data into replicated genes or not (to speed operations on replicated data)
repdata = cluster2[cluster2$NAME %in% replicated,]
norepdata = cluster2[!cluster2$NAME %in% replicated,]
# reduce replicated genes data frame to most variable probe
temp = melt(repdata,id.vars = c("UNIQID","NAME"))
temp = ddply(temp,.(NAME,UNIQID),summarize,iqr=IQR(value))
temp = temp[order(temp$NAME,temp$iqr,decreasing=TRUE),]
temp = temp[!duplicated(temp$NAME),]
repdata$bestprobe = temp$UNIQID[match(repdata$NAME,temp$NAME)]
repdata = repdata[repdata$bestprobe==repdata$UNIQID,]
# re-combine the replicated and non-replicated
cluster2 = rbind(norepdata,repdata[,-which(names(repdata)=="bestprobe")])
```

### 3) Find genes that differentiate between APB and FPB

```{r}
cluster3 = cluster2
# Set rownames to the genes
rownames(cluster3) = cluster3$NAME
cluster3 = cluster3[,3:ncol(cluster3)]
# Remove outlier fetal sample
dedata = cluster3[,-which(names(cluster3)=="5FPBT#5")]
# Create a group variable to hold the age classifier of samples
group = rep("FPB",dim(dedata)[2])
group[grep("APB",names(dedata))] = "APB"
group = factor(group,levels=c("FPB","APB")) 
# Perform differential expression with limma
design=model.matrix(~group)
fit=lmFit(dedata,design)
fit=eBayes(fit)
topgenes = topTable(fit,
                    coef=2,
                    adjust.method="BH",
                    n=Inf)
topgenes <- topgenes[abs(topgenes$logFC)>log2(1.5) & topgenes$adj.P.Val<0.05,]
# Export DE table and the data that went in to it
write.csv(topgenes,
          file = "Microarray_APBvsFPBTcells_FDR0.05_FC1.5_LongAnn.csv",
          row.names=TRUE,
          quote=FALSE)
save(dedata, file = "Tcell_dedata.rda")

# Store this topgenes table for later.
topgenes_T <- topgenes
```

## Monocytes second

### 0) Load in the data and get started

```{r}
# Load most raw form and median scale
cluster1=read.csv("Microarray_annotatedData/17_medGen_longAnn_Mono.csv", header=T,row.names = 1,check.names=FALSE)
```

### 1) Check for outliers:

```{r}
#### CLUSTER FPB AND APB -  cells
cluster2=cluster1[,c(4:13)]
distance = dist(t(cluster2), method="euclidean")
hc <- hclust(distance, method="average")
plot(hc, main = "Cluster Dendrogram", ylab="Height", xlab =NULL)
```

Fetal peripheral blood #5 clusters on its own here and in the T cell analysis.

Additionally, APB#1 and APB#5 cluster with the fetal samples.  These differences are investigated in step 3.

### 2) Select best probes (highest interquartile range) for genes with multiple.

```{r}
cluster2=cluster1[,c(1,2,4:13)]
cluster2$NAME = sapply(as.character(cluster2$NAME),function(x) strsplit(x," ")[[1]][1])

# pick the most variable probe for each replicated gene
replicated = unique(cluster2$NAME[duplicated(cluster2$NAME)])
# separate data into replicated genes or not (to speed operations on replicated data)
repdata = cluster2[cluster2$NAME %in% replicated,]
norepdata = cluster2[!cluster2$NAME %in% replicated,]
# reduce replicated genes data frame to most variable probe
temp = melt(repdata,id.vars = c("UNIQID","NAME"))
temp = ddply(temp,.(NAME,UNIQID),summarize,iqr=IQR(value))
temp = temp[order(temp$NAME,temp$iqr,decreasing=TRUE),]
temp = temp[!duplicated(temp$NAME),]
repdata$bestprobe = temp$UNIQID[match(repdata$NAME,temp$NAME)]
repdata = repdata[repdata$bestprobe==repdata$UNIQID,]
# re-combine the replicated and non-replicated
cluster2 = rbind(norepdata,repdata[,-which(names(repdata)=="bestprobe")])
```

### 3) Investigate Adult samples that cluster with fetal samples

```{r}
cluster3 = cluster2
# Set rownames to the genes
rownames(cluster3) = cluster3$NAME
cluster3 = cluster3[,3:dim(cluster3)[2]]
# Remove outlier fetal sample
dedata = cluster3[,grep("APB",names(cluster3))]
# Create a group variable to hold the age classifier of samples
group = rep(1,dim(dedata)[2])
group[names(dedata) %in% c("2APBmono#5","2APBmono#1")] = 2 
# Perform differential expression with limma
design=model.matrix(~group)
fit=lmFit(dedata,design)
fit=eBayes(fit)
topgenes = topTable(fit,
                    coef=2,
                    adjust.method="BH",
                    p.value=0.05,
                    sort.by="p",
                    resort.by='t',
                    n=100000)
write.csv(topgenes,
          "Microarray_APB_monocyte_Heterogeneity_FDR0.05_noprefilter_IPA_sortbyttest.csv",
          row.names=TRUE,
          quote=FALSE)
```

These genes were used for external pathway analysis which returned pathways, provided in the text, which idicated these samples might have had a viral infection ast the time. The T cell samples were less affected, so these samples were only removed from the monocyte differential expression analysis.  Note: these adult samples were brought back later in training of the bulk immunological age score model.

Differential Expression was run then for monocytes without APB#1, APB#5, and FPB#5.

### 4) Find genes that differentiate between APB and FPB

```{r}
cluster3 = cluster2
# Set rownames to the genes
rownames(cluster3) = cluster3$NAME
cluster3 = cluster3[,3:ncol(cluster3)]
# Remove outlier fetal sample
dedata = cluster3[,-which(names(cluster3) %in% c("1FPBmono#5","2APBmono#5","2APBmono#1"))]
# Create a group variable to hold the age classifier of samples
group = rep("FPB",dim(dedata)[2])
group[grep("APB",names(dedata))] = "APB"
group = factor(group,levels=c("FPB","APB")) 
# Perform differential expression with limma
design=model.matrix(~group)
fit=lmFit(dedata,design)
fit=eBayes(fit)
topgenes = topTable(fit,
                    coef=2,
                    adjust.method="BH",
                    n=Inf)
topgenes <- topgenes[abs(topgenes$logFC)>log2(1.5) & topgenes$adj.P.Val<0.05,]
# Export DE table
write.csv(topgenes,
            file = "Microarray_APBvsFPBmonocytes_FDR0.05_FC1.5_LongAnn.csv",
            row.names=TRUE,
            quote=FALSE)
# Store this topgenes table for later.
topgenes_M <- topgenes

# Export the DEdata with only FPB#5 removed for immunological age score model generation later.
dedata = cluster3[,-which(names(cluster3) %in% c("1FPBmono#5"))]
save(dedata, file = "Monocyte_dedata.rda")
```

# 2. Extract broad fetal versus adult gene signature (Intersect the T and monocyte DE genes)

This is for potential immunological age score marker selection

```{r}
full_Adult <- intersect(rownames(topgenes_T[topgenes_T$logFC>0,]),
                        rownames(topgenes_M[topgenes_M$logFC>0,]))
full_Fetal <- intersect(rownames(topgenes_T[topgenes_T$logFC<0,]),
                        rownames(topgenes_M[topgenes_M$logFC<0,]))
full <- c(full_Adult, full_Fetal)
(filtered <- full[-grep("^XLOC|^LOC|^ENST|ORF|orf|A_19|A_33|A_24", full)])
```

A subset of these genes were validated by qRT-PCR and then used to score additional samples without the need for full microarray analysis.

# 3. Calculate Immunological Age Scores for initial UCSF(SLVR) UCB cohort & controls

## 1) Read in qPCR data & pre-process

### 1. Clean it up

```{r}
chiptMay = read.csv("qRTPCR_data/140527tcell.csv",
                    na.strings = c("NA", "","999"), skip=31, stringsAsFactors = FALSE)
chiptMay$run = "may"
chipmMay = read.csv("qRTPCR_data/140528monoresultsLegit.csv",
                    na.strings = c("NA", "","999"), skip=31, stringsAsFactors = FALSE)
chipmMay = chipmMay[,-which(names(chipmMay)=="Comments")]
chipmMay$run = "may"
chipmMay$Name = sub("M$","Mono",chipmMay$Name)

chiptMarch = read.csv("qRTPCR_data/140304 T cell layering.csv",
                      na.strings = c("NA", ""), skip=11, stringsAsFactors = FALSE)
chiptMarch = chiptMarch[,!names(chiptMarch) %in% c("In.Range","Out.Range","Peak.Ratio")]
chiptMarch$run = "march"
chipmMarch = read.csv("qRTPCR_data/140304 Mono layering.csv",
                      na.strings = c("NA", ""), skip=11, stringsAsFactors = FALSE)
chipmMarch = chipmMarch[,!names(chipmMarch) %in% c("In.Range","Out.Range","Peak.Ratio")]
chipmMarch$run = "march"

d = rbind(chiptMay,chiptMarch,chipmMay,chipmMarch)

names(d)[names(d)=="Name"] = "sample"
names(d)[names(d)=="Name.1"] = "gene"
names(d)[names(d)=="Value"] = "ct"

# get rid of whitespace in gene names
d$gene = gsub(" ","",d$gene)

# get rid of unused sample or assay inlets
d = d[d$sample != "BLANK" & d$gene != "BLANK" & !is.na(d$gene) & !is.na(d$sample),]

# SLVR035 in May should be labeled SLVR135
d$sample[d$run=="may"] = sub("SLVR035","SLVR135",d$sample[d$run=="may"])

# expand sample pheno data
d$id = sapply(d$sample,function (x) strsplit(x,".",fixed=TRUE)[[1]][1])
d$rep = as.numeric(sapply(d$sample,function (x) strsplit(x,".",fixed=TRUE)[[1]][2]))
d$cell = sapply(d$sample,function (x) strsplit(x,".",fixed=TRUE)[[1]][3])
```

### 2. Filter ct data (simply set to NA)

```{r}
d$ct[d$Quality<0.5] = NA
# Across all genes, any Ct > pre-specified cutoff (e.g. 28), is assigned NA.
d$ct[d$ct>27] = NA
```

### 3. Trim down to just the signature genes

```{r}
siggenes <- as.character(t(read.csv("qRTPCR_data/Signature_Genes.csv")))
d = d[d$gene %in% siggenes,]
```

### 4. Trim samples with fewer than 3 successful replicates

```{r}
### Trim replicates (greater than 80% failed reactions)

# remove shared samples and ctrl samples from may run
d = d[-which(grepl("ctrl|SLVR022|SLVR031|SLVR135",d$sample) & d$run=="may"),]

# identify likely bad replicates by fraction failed
d = ddply(d,.(sample,cell,run),transform,remove = sum(is.na(ct))/length(ct)>=0.8)

# list s that will be removed
unique(paste(d$sample[d$remove],d$run[d$remove]))
ddply(d[d$remove,],.(id,cell,run),summarize,number.reps = length(unique(rep)))
ddply(d[d$remove,],.(id,cell,rep),summarize,fraction.failed = sum(is.na(ct))/length(ct))

# remove likely bad runs by fraction failed
if (sum(d$remove)>0){
  d = d[-which(d$remove),]
}

# identify and remove samples that are outliers (outside 0.01 tails on a normal distribution when looking at data for a given gene and cell type, across all cords. Outside 0.01 tails means total of 2*0.01 = 0.02 or 2% of the data is removed, leaving 98%)
d = ddply(d,.(cell,gene),transform,ctoutliersremoved = outliersNA(ct,0.01))
d$ct = d$ctoutliersremoved
```

```{r}
### Trim samples with fewer than 3 replicates
# remove subjects entirely for a given cell type that don't have at least 3 wells of data
d = ddply(d,.(id,cell,gene),transform,remove = length(unique(rep))<3)

# list additional samples that will be removed
unique(paste(d$id[d$remove]))

# remove these additional samples
if (sum(d$remove)>0) {
  d = d[-which(d$remove),]
}
```

### 5. Trim genes

```{r}
### get rid of likely bad genes with greater than 80% reactions failed
d = ddply(d,.(gene,cell,run),transform,remove = sum(is.na(ct))/length(ct)>=0.8)
# list genes that will be removed
unique(paste(d$gene[d$remove],d$cell[d$remove],d$run[d$remove]))
# remove these likely bad genes
if (sum(d$remove)>0) {
  d = d[-which(d$remove),]
}
```

```{r}
# remove genes that are now not shared across both runs within a given cell type
d = ddply(d,.(gene,cell),transform,remove = length(unique(run))<2)
# list genes that will be removed
unique(paste(d$gene[d$remove],d$cell[d$remove]))
# remove likely bad genes
if (sum(d$remove)>0) {
  d = d[-which(d$remove),]
}
```

```{r}
# get rid of genes that have average Ct > than a certain amount, here 25
d = ddply(d,.(gene),transform,remove = mean(ct,na.rm=TRUE) > 25)
# list genes that will be removed
unique(paste(d$gene[d$remove],d$cell[d$remove]))
if (sum(d$remove)>0) {
  d = d[-which(d$remove),]
}
```

### 6. Normalize between runs based on mean gene Ct values
```{r}
# compute gene-specific normalization factor, which is the mean of the samples across a run within a cell type
d = ddply(d,.(gene,run,cell),transform,NF = mean(ct,na.rm=TRUE))
# apply normalization
d$ctnorm = d$ct - d$NF
```

### 7. Convert normalized ct to relative expression

```{r}
# On a per-gene-run-cell basis, take the highest observed Ct, add 0.1, and assign that to all NA reactions for that gene.
d = ddply(d,.(gene,run,cell),transform,max = max(ctnorm,na.rm=TRUE))
d$ctnorm[is.na(d$ctnorm)] = d$max[is.na(d$ctnorm)] + 0.1

# Perform Transformation
d$log2exp = -1*d$ctnorm
d = ddply(d,.(gene),transform,log2exp.scaled = log2exp - min(log2exp,na.rm=TRUE))
```

## 2) Calculate scoring weights based on microarray PCA of signature genes suviving quality control

### 1. Generate loadings for T cells

```{r}
# load array data in which most variable probe has been retained and 5FPB#5 already removed; the data is loaded into dedata
load("Tcell_dedata.rda")
dedata = dedata[,grep("FPB|APB",names(dedata))]

# reduce array data to genes of interest
dedata = dedata[rownames(dedata) %in% d$gene[d$cell=="T"],]

# extract pheno data for pca plot
group = rep("FPB",dim(dedata)[2])
group[grep("APB",names(dedata))] = "APB"
names(group) = names(dedata)
pcdat = prcomp(x=t(dedata),scale.=TRUE)

# get PC1 loadings to make signature
Tsig = data.frame(pcdat$rotation[,1:2])[,1]
names(Tsig) = rownames(data.frame(pcdat$rotation[,1:2]))

# what % of variation in array data is explained by PC1
(pcdat$sdev^2)[1]/sum(pcdat$sdev^2)
```

```{r}
#Loadings
Tsig
```

#### Plot the PCA (Figure S2)

```{r, fig.width=2.33, fig.height = 2.4}
# Gather data
pc <- data.frame(
  pcdat$x[,1:2],
  group=group[match(rownames(pcdat$x),names(group))],
  sample=rownames(pcdat$x))
# get loadings
loadings <- data.frame(pcdat$rotation[,1:2])
# amplify loadings so we can see them on the plot
loadings <- loadings * 30
loadings$gene <- rownames(loadings)
# Make a pretty plot
(p = ggplot(data=pc,
           aes(PC1, PC2)) +
  geom_point(aes(fill=group),
             color="black", size=3, shape = 21) +
  scale_fill_manual(values=dittoColors()[c(1,3)]) +
  geom_segment(data=loadings,
               aes(x=0, y=0, xend=PC1, yend=PC2),
               arrow=arrow(length=unit(0.2,"cm")),
               alpha=0.25) +
  geom_text_repel(data=loadings[loadings$PC1<0,],
            direction = "both", xlim = c(NA, -7),
            aes(x=PC1, y=PC2, label=gene, fontface = 2),
            force = 5, seed = 3, min.segment.length = 0.05,
            size=2, segment.size = 0.2,
            box.padding = 0.02) +
  geom_text_repel(data=loadings[loadings$PC1>0,],
            direction = "both", xlim = c(7, NA),
            aes(x=PC1, y=PC2, label=gene, fontface = 2),
            force = 3, seed = 5, min.segment.length = 0.05,
            size=2, segment.size = 0.2,
            box.padding = 0.1) +
  xlab(paste0("PC1 (",
              round(100*(pcdat$sdev^2)[1]/sum(pcdat$sdev^2), digits = 1),
              "%)")) +
  ylab(paste0("PC2 (",
              round(100*(pcdat$sdev^2)[2]/sum(pcdat$sdev^2), digits = 1),
              "%)")) +
  theme_bw() + ggtitle(NULL, subtitle = "T cells") +
  theme(axis.title.x= element_text(size = 8),
        axis.text.x= element_text(size = 8, face = "bold"),
        axis.title.y= element_text(size = 8),
        axis.text.y= element_text(size = 8),
        plot.subtitle= element_text(size = 9)) +
  coord_cartesian(clip = "off", xlim = c(-15,13)) +
  theme(legend.position = "none"))
# p = p + geom_text(aes(label=sample))
pdf("Microarray-qPCR-Figs/PCA_T_Array.pdf", w = 2.33, h = 2.4)
p
dev.off()
```

#### Plot the loadings (Figure S2)

```{r, fig.height=2.5, fig.width=1.5}
Tdat <- data.frame(
    weights = Tsig,
    genes = factor(names(Tsig), levels = names(Tsig[order(Tsig)])),
    sign = Tsig>0)
(p <- ggplot(Tdat, aes(x = genes, y = weights, fill = sign)) +
  geom_col(color = "black", width = 0.7) + coord_flip() + theme_classic() +
  scale_fill_manual(values = c("blue3","red3"), guide = FALSE) +
  geom_hline(yintercept = 0) +
  scale_y_continuous(breaks = c(-0.2,0,0.2)) +
  theme(axis.text.x= element_text(size = 7),
        axis.text.y= element_text(size = 6),
        axis.title.x= element_text(size = 7),
        plot.title = element_text(size = 9)) +
  xlab(NULL))
pdf("Microarray-qPCR-Figs/Tsig_weights.pdf", w=1.5, h=2.5)
p
dev.off()
```

### 2. Generate loadings for Monocytes

```{r}
# load array data in which most variable probe has been retained and 5FPB#5 already removed; the data is loaded into dedata
load("Monocyte_dedata.rda")
dedata = dedata[,grep("FPB|APB",names(dedata))]

# reduce array data to genes of interest
dedata = dedata[rownames(dedata) %in% d$gene[d$cell=="Mono"],]

# extract pheno data for pca plot
group = rep("FPB",dim(dedata)[2])
group[grep("APB",names(dedata))] = "APB"
names(group) = names(dedata)
pcdat = prcomp(x=t(dedata),scale.=TRUE)

# get PC1 loadings to make signature
Msig = data.frame(pcdat$rotation[,1:2])[,1]
names(Msig) = rownames(data.frame(pcdat$rotation[,1:2]))

# what % of variation in array data is explained by PC1
(pcdat$sdev^2)[1]/sum(pcdat$sdev^2)
```

#### Plot the PCA (Figure S2)

```{r, fig.width=2.4, fig.height=2.4}
# Gather data
pc <- data.frame(pcdat$x[,1:2],group=group[match(rownames(pcdat$x),names(group))],sample=rownames(pcdat$x))
# get loadings
loadings <- data.frame(pcdat$rotation[,1:2])
# amplify loadings so we can see them on the plot
loadings <- loadings * 35
loadings$gene <- rownames(loadings)
# Make a pretty plot
(p <- ggplot(data=pc,
           aes(PC1, PC2)) +
  geom_point(aes(fill=group),
             color="black", size=3, shape = 21) +
  scale_fill_manual(values=dittoColors()[c(1,3)]) +
  geom_segment(data=loadings,
               aes(x=0, y=0, xend=PC1, yend=PC2),
               arrow=arrow(length=unit(0.2,"cm")),
               alpha=0.25) +
  geom_text_repel(data=loadings[loadings$PC1<0,],
            direction = "both", xlim = c(NA, -7),
            aes(x=PC1, y=PC2, label=gene, fontface = 2),
            force = 3, seed = 5, min.segment.length = 0.05,
            size=2, segment.size = 0.2,
            box.padding = 0.01,
            max.iter = 1e5) +
  geom_text_repel(data=loadings[loadings$PC1>0,],
            direction = "both", xlim = c(7, NA),
            aes(x=PC1, y=PC2, label=gene, fontface = 2),
            force = 3, seed = 5, min.segment.length = 0.05,
            size=2, segment.size = 0.2,
            box.padding = 0.01,
            max.iter = 1e5) +
  xlab(paste0("PC1 (",
              round(100*(pcdat$sdev^2)[1]/sum(pcdat$sdev^2), digits = 1),
              "%)")) +
  ylab(paste0("PC2 (",
              round(100*(pcdat$sdev^2)[2]/sum(pcdat$sdev^2), digits = 1),
              "%)")) +
  theme_bw() + ggtitle(NULL, subtitle = "Monocytes") +
  theme(axis.title.x= element_text(size = 8),
        axis.text.x= element_text(size = 8, face = "bold"),
        axis.title.y= element_text(size = 8),
        axis.text.y= element_text(size = 8),
        plot.subtitle= element_text(size = 9)) +
  coord_cartesian(clip = "off", xlim = c(-13,13)) +
  theme(legend.position = "none"))
# p = p + geom_text(aes(label=sample))
pdf("Microarray-qPCR-Figs/PCA_Mono_Array.pdf", w = 2.4, h = 2.4)
p
dev.off()
```

#### Plot the loadings (Figure S2)

```{r, fig.width=1.5, fig.height=2.5}
Mdat <- data.frame(
    weights = Msig,
    genes = factor(names(Msig), levels = names(Msig[order(Msig)])),
    sign = Msig>0)
(p <- ggplot(Mdat, aes(x = genes, y = weights, fill = sign)) +
  geom_col(color = "black", width = 0.7) + coord_flip() + theme_classic() +
  scale_fill_manual(values = c("blue3","red3"), guide = FALSE) +
  geom_hline(yintercept = 0) +
  scale_y_continuous(breaks = c(-0.2,0,0.2)) +
  theme(axis.text.x= element_text(size = 7),
        axis.text.y= element_text(size = 6),
        axis.title.x= element_text(size = 7),
        plot.title = element_text(size = 9)) +
  xlab(NULL))
pdf("Microarray-qPCR-Figs/Msig_weights.pdf", w=1.5, h=2.5)
p
dev.off()
```

## 3) Apply the signature to Fetal, UCB, and adult qPCR or microarray samples

### 1. To UCB qPCR

```{r}
# remember, d is the qPCR data on the cord blood samples
dt = d[d$cell == "T" & d$gene %in% names(Tsig),]
dt = ddply(dt,.(id,gene,run),summarize,log2exp = mean(log2exp,na.rm=TRUE))
dt = ddply(dt,.(gene),transform,log2exp.stand = scale(log2exp))

dm = d[d$cell == "Mono" & d$gene %in% names(Msig),]
dm = ddply(dm,.(id,gene,run),summarize,log2exp = mean(log2exp,na.rm=TRUE))
dm = ddply(dm,.(gene),transform,log2exp.stand = scale(log2exp))

dt$coef = Tsig[match(dt$gene,names(Tsig))]
dm$coef = Msig[match(dm$gene,names(Msig))]

dt = ddply(dt,.(id,run),summarize,Tsig = sum(log2exp.stand * coef))
dm = ddply(dm,.(id,run),summarize,Msig = sum(log2exp.stand * coef))
```

### 2. To fetal & adult qPCR

```{r}
load("qRTPCR_data/Controls/dconJune.RData")
# the data from the preceeding line loads in a data frame called dconJune, which was created from the C1 data in a similar method to the above, using the script analysis_nonC1samples_updated.R

# reduce control data to genes from SILVER runs
all(unique(d$gene) %in% unique(dconJune$gene))
dconJune = dconJune[dconJune$gene %in% unique(d$gene),]

### T Cells ###

# Separate out t cell data
dcont = dconJune[dconJune$cell == "t" & dconJune$gene %in% names(Tsig),]
# average replicates
dcont = ddply(dcont,.(tissue,subject,gene,cord),summarize,log2exp = mean(log2exp,na.rm=TRUE))
# standardize each gene (z score)
dcont = ddply(dcont,.(gene),transform,log2exp.stand = scale(log2exp))

# get gene weights from signature
dcont$coef = Tsig[match(dcont$gene,names(Tsig))]

# compute signature scores
dcont = ddply(dcont,.(tissue,subject,cord),summarize,Tsig = sum(log2exp.stand * coef))

### Monocytes ###

# Separate out t cell data
dconm = dconJune[dconJune$cell == "m" & dconJune$gene %in% names(Msig),]
# average replicates
dconm = ddply(dconm,.(tissue,subject,gene,cord),summarize,log2exp = mean(log2exp,na.rm=TRUE))
# standardize each gene (z score)
dconm = ddply(dconm,.(gene),transform,log2exp.stand = scale(log2exp))

# get gene weights from signature
dconm$coef = Msig[match(dconm$gene,names(Msig))]

# compute signature scores
dconm = ddply(dconm,.(tissue,subject,cord),summarize,Msig = sum(log2exp.stand * coef))
```

### 3. To previous microarrays

This function takes in a GEO ExpressionSet, then trims to (max variance probe for) each signature gene, 

```{r}
eset_to_ImmAgeScores <- function(eset = Monocyte_eset, loadings = Msig) {
  # INPUT: Takes in an ExpressionSet (eset) obtained from GEO, and a named
    # vector gene loadings for Immunological Age Scoring (loadings; values
    # named by gene_symbols)
  # VALUE: Vector of Immunological Age Scores for each sample in the eset
  # DETAILS: It subsets the loadings to genes probed in the expression dataset,
    # and utilizes expression data for the probes, of each gene in loadings,
    # with the highest variance as measured by interquartile range,
    # then calculates Immunological Age Scores by summing the product of the
    # loading for each gene * the relative (z-score) expression of each sample
    # for each gene.

  # get all the array probe ids that match the genes in th Scoring loadings
  probeids <- as.character(fData(eset)$ID)	
  probelist <- sapply(names(loadings),function(x) probeids[which(fData(eset)$GENE_SYMBOL==x)])
  
  # retain only most variable probe
  mostvar <- function(theprobes) {
  	vars <- sapply(theprobes,function(x) IQR(exprs(eset)[which(fData(eset)$ID==x),]))
  	return(theprobes[which.max(vars)])
  }
  probelist <- llply(probelist,mostvar)
  
  # convert array probe ids into a single vector,
  # which will have any genes not probed in the dataset removed.
  probes_use = NULL
  for (curitem in seq_along(probelist)) {
  	probes_use = c(probes_use,probelist[[curitem]])
  }
  if (length(probes_use)!=length(probelist)) {
    not_found_inds <- sapply(seq_along(probelist),
                             function(X) length(probelist[[X]])==0)
    warning(paste0(names(probelist)[not_found_inds], collapse = ", "),
            " not found in the dataset.\n")
    probelist <- probelist[!not_found_inds]
    loadings <- loadings[!not_found_inds]
  }
  
  # Obtain relevant expression data
  dedata <- exprs(eset)[probes_use,]
  rownames(dedata) <- names(probelist)
  
  # Calculate scores based on scaled (relative!) expression within the dataset
  scaled_data <- t(scale(t(dedata)))
  Scores <- apply(scaled_data*loadings,2,sum)
}
```

#### Mold et al., Science, 2010 T cell data.

```{r}
# Obtain data and rename samples by their titles
Tcell_eset <- getGEO(GEO = "GSE25087")[[1]]
colnames(Tcell_eset) <- Tcell_eset$title

# Trim to naive T cells (dataset also contains CD25+ Tregs)
Tcell_eset <- Tcell_eset[,grep("Naïve",Tcell_eset$title)]

# Correct the name of the datasets gene_symbol feature data
fData(Tcell_eset)$GENE_SYMBOL = fData(Tcell_eset)$'Gene Symbol'
```

```{r}
# Score
(Tmicro <- eset_to_ImmAgeScores(eset = Tcell_eset, loadings = Tsig))
```

#### Krow-Lucal et al., Blood, 2014 Monocyte data.

```{r}
# Obtain data and rename samples by their titles
Monocyte_eset <- getGEO(GEO = "GSE54818")[[1]]
colnames(Monocyte_eset) <- Monocyte_eset$title

# Trim to unstimulated cells (dataset also contains IFNg stimulations)
Monocyte_eset <- Monocyte_eset[,grep("Unstim",Monocyte_eset$title)]
```

```{r}
# Score
(Mmicro <- eset_to_ImmAgeScores(eset = Monocyte_eset, loadings = Msig))
```

## Combine all of the scoring and annotations into a single data.frame for plotting

```{r}
#Create data frame
tdata <- data.frame(
  sample.name = c(names(Tmicro), # Tcell microarray
                  dt$id, # UCB qPCR
                  paste(dcont$tissue,dcont$subject)), # control qPCR
  signature=c(Tmicro,
              dt$Tsig,
              dcont$Tsig),
  age = c(rep("fetal", 3), #Tmicro has three fetal, then 3 adult
          rep("adult", 3),
          rep("UCB", 29), # UCB qPCR data has 29 samples after all trimming
          rep("adult", 3), # Fetal and adult control qPCR has 3 adult, then 3 fetal
          rep("fetal", 3)),
  tissue=c(rep("FmLN",3), # mLN = mesenteric lymph node
           rep("APB", 3),
           rep("UCB",dim(dt)[1]),
           toupper(dcont$tissue)),
  run = c(rep("array",length(Tmicro)),
          paste("qPCR",dt$run),
          paste("qPCR June",dcont$cord)),
  platform = c(rep("array",length(Tmicro)),
               rep("qPCR",length(dt$run)),
               rep("qPCR",length(dcont$cord))),
  celltype = "T cells")
mdata <- data.frame(
  sample.name = c(names(Mmicro), # Monocyte microarray
                  dm$id, # UCB qPCR
                  paste(dconm$tissue,dconm$subject)), # fetal and adult control qPCR
  signature=c(Mmicro,
              dm$Msig,
              dconm$Msig),
  age = c(rep("adult", 4), #Mmicro has four adult, then 4 fetal
          rep("fetal", 4),
          rep("UCB", 27), # UCB qPCR data has 27 samples after all trimming
          rep("adult", 3), # Fetal and adult control qPCR has 3 adult, then 3 fetal
          rep("fetal", 3)),
  tissue=c(rep("ABM",4), # Mmicro used bone marrow tissues.
           rep("FBM", 4),
           rep("UCB",dim(dm)[1]),
           toupper(dconm$tissue)),
  run = c(rep("array",length(Mmicro)),
          paste("qPCR",dm$run),
          paste("qPCR June",dconm$cord)),
  platform = c(rep("array",length(Mmicro)),
               rep("qPCR",length(dm$run)),
               rep("qPCR",length(dconm$cord))),
  celltype = "Monocytes")

# Change incorrectly labeled FLN from control qPCR data to FS (fetal spleen)
tdata$tissue <- as.character(tdata$tissue)
tdata$tissue[tdata$tissue=="FLN"] <- "FS"

# Combine T and M data
alldata <- rbind(tdata,mdata)

# Change order of how tissues will be plotted.
alldata$tissue <- factor(
  alldata$tissue,
  levels=c("FPB","FS","FmLN","FBM","UCB","APB","ABM"))

# Add line with both age and celltype
alldata$agetype <- paste(alldata$age, alldata$celltype, sep = "\n")
# Change order of how ages will be plotted.
alldata$agetype <- factor(
  alldata$agetype,
  levels = c("fetal\nT cells", "adult\nT cells", "UCB\nT cells",
             "fetal\nMonocytes", "adult\nMonocytes", "UCB\nMonocytes"))
alldata$age <- factor(
  alldata$age,
  levels=c("fetal","adult","UCB"))

## Save!
save(alldata, file = "Microarray-qPCR-Figs/sigscoring_withControls_alldata.rdata")
```

Also calculate summary statistics per group for adding to plots

```{r}
# Specifically, we want standard deviation and standard error, but this also calculates a bunch more.
summaryStats <- summarySE(
  alldata, measurevar="signature",
  groupvars=c("age","platform","tissue","celltype"))
summaryStats$agetype <- paste(summaryStats$age, summaryStats$celltype, sep = "\n")
```

# 4. Plot scores

## UCSF/SLVR qPCR with fetal and adult qPCR controls (Figure 2)

```{r, fig.width=3, fig.height=3}
### Fetal and Adult Scores

(p <- ggplot(
  alldata[
    alldata$platform=="qPCR" & alldata$age %in% c("fetal", "adult"),],
  aes(x = agetype, y = signature, fill = age),
  color = "black") +
  # scores
  geom_jitter(position=position_jitter(w=0.2,h=0), shape = 21) + 
  # SEM bars
   geom_errorbar(
    data = summaryStats[
      summaryStats$platform=="qPCR" & summaryStats$age %in% c("fetal", "adult"),],
    aes(x = agetype, ymin = signature - se, ymax = signature + se),
    width = 0.3, color = "black") +
  # mean bar
  geom_errorbar(
    data = summaryStats[
      summaryStats$platform=="qPCR" & summaryStats$age %in% c("fetal", "adult"),],
    aes(x = agetype, ymin = signature, ymax = signature),
    width = 0.5, color = "black") +
  # Plot theme and titles
  theme_classic() +
  theme(legend.position = "none") +
  ylab("Immunological Age Score") +
  xlab(NULL) +
  ggtitle(NULL, subtitle = " Fetal and Adult\nassessed by qRT-PCR") +
  theme(axis.text.x= element_text(angle=45, hjust=1, vjust=1),
        axis.title.y= element_text(size = 10),
        plot.subtitle = element_text(size = 11)) +
  coord_cartesian(ylim = c(-4, 4)) +
  scale_fill_manual(values = dittoColors()[c(3,1)]))

pdf("Microarray-qPCR-Figs/SigScoring_FandA_qPCR.pdf",w=3,h=3)
p
dev.off()
```

```{r, fig.width=2.1, fig.height=3}
# Create a cord data for adding violins without rearranging the age order
corddata <- alldata[alldata$platform=="qPCR" & alldata$age=="UCB",]

### Classical Monocyte Scores
(p <- ggplot(
  corddata[corddata$platform=="qPCR",],
  aes(x = agetype, y = signature, fill = age),
  color = "black") +
  # scores
  geom_violin(alpha = 0.75) +
  geom_jitter(position=position_jitter(w=0.2,h=0), shape = 21) + 
  # SEM
  geom_errorbar(
    data = summaryStats[
      summaryStats$platform=="qPCR" & summaryStats$age=="UCB",],
    aes(x = agetype, ymin = signature - se, ymax = signature + se),
    width = 0.3, color = "black") +
  # mean
  geom_errorbar(
    data = summaryStats[
      summaryStats$platform=="qPCR" & summaryStats$age=="UCB",],
    aes(x = agetype, ymin = signature, ymax = signature),
    width = 0.5, color = "black") +
  # Plot theme and titles
  theme_classic() +
  theme(legend.position = "none") +
  ylab("Immunological Age Score") +
  xlab(NULL) + ggtitle(NULL, subtitle = " Newborn UCB\nassessed by qRT-PCR") +
  theme(axis.text.x= element_text(angle=45, hjust=1, vjust=1),
        axis.title.y= element_text(size = 10),
        plot.subtitle = element_text(size = 11)) +
  coord_cartesian(ylim = c(-4, 4)) +
  scale_fill_manual(values = dittoColors()[c(2)]))

pdf("Microarray-qPCR-Figs/SigScoring_UCB_qPCR.pdf",w=2.1,h=3)
p
dev.off()
```

```{r, fig.width=2.7, fig.height=2.7}
# Correlation of T cell versus monocyte scores

corddata.bysample <- 
    dplyr::inner_join(
        alldata[alldata$celltype=="T cells" & alldata$age=="UCB",c(1,2)],
        alldata[alldata$celltype=="Monocytes" & alldata$age=="UCB",c(1,2)],
        by = "sample.name")
names(corddata.bysample) <- c("sample.name", "Tsig", "Msig")
# Calculate Correlation statistics
cor <- cor.test(corddata.bysample$Tsig,corddata.bysample$Msig)
# Make the plot
(p <- ggplot(corddata.bysample, aes(x = Tsig, y = Msig)) +
  geom_smooth(method = "lm", se = FALSE, color = "black", size = 1.5) +
  geom_point(fill = dittoColors()[2], color = "black", shape = 21) +
  theme_bw() +
  theme(legend.position = "none") +
  ylab("Monocytes") +
  xlab("T cells") +
  ggtitle(NULL, subtitle = "Correlation of scores") +
  geom_text(
      data = data.frame(
          x = c(0.6,0.6),
          y = c(-1.6, -2.1),
          label = c(
              paste0("r = ", signif(cor$estimate,2)),
              paste0("p = ", signif(cor$p.value,2)))),
      aes(x = x, y = y, label = label))+
  theme(axis.title.x= element_text(size = 10),
        axis.title.y= element_text(size = 10),
        plot.subtitle = element_text(size = 11)) +
  coord_cartesian(clip = "off"))
pdf("Microarray-qPCR-Figs/SigScoring_Correlation_qPCR.pdf",w=2.7,h=2.7)
p
dev.off()
```

## Previous microarrays (Figure S2)

```{r, fig.height=2, fig.width=1.5}
# MicroArray:
(p <- ggplot(alldata[alldata$platform=="array" & alldata$celltype=="Monocytes",],
       aes(x = age, y = signature, fill = age)) + 
  # geom_boxplot(outlier.shape=NA,fill=NA) +
  geom_jitter(position=position_jitter(w=0.2,h=0), shape = 21) + 
  geom_errorbar(data = summaryStats[summaryStats$platform=="array" & summaryStats$celltype=="Monocytes",],
                aes(x = age, ymin = signature - se, ymax = signature + se),
                width = 0.3,
                color = "black") +
  geom_errorbar(data = summaryStats[summaryStats$platform=="array" & summaryStats$celltype=="Monocytes",],
                aes(x = age, ymin = signature, ymax = signature),
                width = 0.5,
                color = "black") +
  theme_classic() +
  theme(legend.position = "none") +
  ylab("Microarray-based\nImmunological Age Score") +
  xlab(NULL) + ggtitle(NULL, subtitle = "Monocytes\nmicroarray") +
  theme(axis.text.x= element_text(angle=45, hjust=1, vjust=1),
        axis.title.y= element_text(size = 8),
        plot.subtitle = element_text(size = 9)) +
  scale_fill_manual(values = dittoColors()[c(3,1)]))
pdf("Microarray-qPCR-Figs/SigScoring_Monocytes_array.pdf",w=1.5,h=2)
p
dev.off()

(p <- ggplot(alldata[alldata$platform=="array" & alldata$celltype=="T cells",],
       aes(x = age, y = signature, fill = age)) + 
  # geom_boxplot(outlier.shape=NA,fill=NA) +
  geom_jitter(position=position_jitter(w=0.2,h=0), shape = 21) + 
  geom_errorbar(data = summaryStats[summaryStats$platform=="array" & summaryStats$celltype=="T cells",],
                aes(x = age, ymin = signature - se, ymax = signature + se),
                width = 0.3,
                color = "black") +
  geom_errorbar(data = summaryStats[summaryStats$platform=="array" & summaryStats$celltype=="T cells",],
                aes(x = age, ymin = signature, ymax = signature),
                width = 0.5,
                color = "black") +
  theme_classic() +
  theme(legend.position = "none") +
  ylab("Microarray-based\nImmunological Age Score") +
  xlab(NULL) + ggtitle(NULL, subtitle = "  T cells\nmicroarray") +
  theme(axis.text.x= element_text(angle=45, hjust=1, vjust=1),
        axis.title.y= element_text(size = 8),
        plot.subtitle = element_text(size = 9)) +
  scale_fill_manual(values = dittoColors()[c(3,1)]))
pdf("Microarray-qPCR-Figs/SigScoring_Tcells_array.pdf",w=1.5,h=2)
p
dev.off()
```
