---
title: "Naive T cell scRNA-seq import and analysis"
author: "Dan Bunis"
date: "9/28/2020"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 6, fig.width = 7, warning = F)
library(MAST)
# BiocManager::install("dtm2451/dittoSeq@a3bfe2b")
library(dittoSeq)
library(caret)
library(ranger)
# devtools::install_github("vqv/ggbiplot")
# devtools::install_github("enriquea/feser")
library(feseR)
library(pROC)
# install.packages("reticulate")
library(reticulate)
# use_python("/anaconda3/envs/r-reticulate/bin/python")
# # Run this in Terminal
# # conda create -y -c conda-forge -n r-reticulate umap-learn=0.3.10
use_condaenv("seurat", required = TRUE)
# devtools::install_version(package = "adehabitat", version = "1.8.20")
# devtools::install_version(package = "SDMTools", version = "1.1-221.1")
# devtools::install_version(package = "Seurat", version = "3.0.2")
library(Seurat)
```

# 1. Import CellRanger output and add some basic metadata.

```{r}
Tcells <- CreateSeuratObject(Read10X("Tcells/cellranger_Raw/"))

#Count and add percent.mito meta
Tcells[["percent.mito"]] <- PercentageFeatureSet(Tcells, pattern = "^MT-")
#Count and add percent.ribo meta
Tcells[["percent.ribo"]] <- PercentageFeatureSet(Tcells, pattern = "^RPS|^RPL")
```

# 2. Filter Cells based on #genes, #UMI, %mitochondrial

Subset to only cells:

- with at least 750 genes
- with at least 1500 UMIs
- with less than 5% mitochondrial UMIs

```{r}
# FILTER by minimum #genes
Tcells.cut <- subset(Tcells, subset = nFeature_RNA > 750)
# FILTER by #UMI
Tcells.cut <- subset(Tcells.cut, subset = nCount_RNA > 1500)
# FILTER by maximum percent.mito
Tcells.cut <- subset(Tcells.cut, subset = percent.mito < 5)
Tcells
Tcells.cut
```

# 3. Import Demuxlet information into Seurat.

The data wass generated with three separate 10X lanes, labeled here as CD4, CD4-8, and CD8

```{r}
Tcells.cut <- importDemux(
  Tcells.cut,
  demuxlet.best = c("Tcells/Demuxlet/CD4.best",
                    "Tcells/Demuxlet/CD4-8.best",
                    "Tcells/Demuxlet/CD8.best"),
  lane.names = c("CD4","CD4-8","CD8"))

#Remove the "CD4_" at the start of all my sample names due to coming from RNAseq data with those names.
Tcells.cut[["Sample"]] <- sapply(
    meta("Sample",Tcells.cut),
    function(X) strsplit(X, split = "CD4_")[[1]][2])
```

## Assessmeent of Demuxlet success (before trimming out doublets)

```{r}
demux.SNP.summary(
    Tcells.cut,
    plots = c("jitter","vlnplot","boxplot"),
    boxplot.color = "white",
    boxplot.fill = F,
    add.line = 50)
demux.calls.summary(Tcells.cut, singlets.only = FALSE)
```

The average SNP count per cell given above is quite high compared to the minimum SNPs required for making highly informed, accurate calls.

# 4. Filter on Demuxlet Singlets, and correct lanes, then add more metadata

```{r}
Tcells <- subset(Tcells.cut, subset = demux.doublet.call == "SNG")
table(meta("Lane", Tcells), meta("Sample", Tcells))
```

We can also see, is that there are very few cells called as samples that were not included in their lanes.  (The 1 in APB4, Lane3).  We'll remove this cell just a bit later

```{r}
rm(Tcells.cut)
```

## Add some extra meta datas:

```{r}
# Add Age
Tcells@meta.data$age <- "unknown"
Tcells@meta.data$age[grep("FS",Tcells$Sample)] <- "fetal"
Tcells@meta.data$age[grep("UCB",Tcells$Sample)] <- "cord"
Tcells@meta.data$age[grep("APB",Tcells$Sample)] <- "adult"

# Add Tcelltype
    # This is dependent on both sample-assignment and 10X-lane, so I will need to add how samples were added to each lane
    samples.4 <- c("FS3", "FS4", "FS5", "UCB1", "UCB2", "UCB5", "APB1", "APB2", "APB4", "APB5")
    samples.8 <- c("FS1", "FS5", "UCB2", "UCB3", "UCB4", "UCB5", "APB2", "APB3", "APB5")
    samples.48.4 <- c("FS1", "FS2", "UCB4", "UCB3", "APB3") 
    samples.48.8 <- c("FS3", "UCB1", "APB1", "APB4")
    samples.48 <- c(samples.48.4,samples.48.8)
#For all the cells in each Lane, add the Tcelltype to a "Tcelltype" metadata iff the cell is assigned to a sample from its lane, and a singlet
Tcells@meta.data$Tcelltype <- NA
Tcells@meta.data$Tcelltype[(Tcells$Sample%in%samples.4) & (Tcells$Lane=="CD4")] <- "CD4"
Tcells@meta.data$Tcelltype[(Tcells$Sample%in%samples.48.4) & (Tcells$Lane=="CD4-8")] <- "CD4"
Tcells@meta.data$Tcelltype[(Tcells$Sample%in%samples.8) & (Tcells$Lane=="CD8")] <- "CD8"
Tcells@meta.data$Tcelltype[(Tcells$Sample%in%samples.48.8) & (Tcells$Lane=="CD4-8")] <- "CD8"

#Add a Tage metadata so that I can plot things with the ages and CD4s vs CD8s separately:
Tcells@meta.data$Tage <- NA
Tcells@meta.data$Tage[Tcells$Tcelltype=="CD4"&!is.na(Tcells$Tcelltype)] <- paste0("4-",Tcells$age[Tcells$Tcelltype=="CD4"&!is.na(Tcells$Tcelltype)])
Tcells@meta.data$Tage[Tcells$Tcelltype=="CD8"&!is.na(Tcells$Tcelltype)] <- paste0("8-",Tcells$age[Tcells$Tcelltype=="CD8"&!is.na(Tcells$Tcelltype)])
```

## Subset based on cells being called to a sample from their lane:

```{r}
# The NA needs to be changed to a value first for Seurat's subset function.
sum(is.na(Tcells$Tage))
Tcells@meta.data$Tage[is.na(Tcells$Tage)] <- "0"
sum(Tcells$Tage=="0")
Tcells <- subset(Tcells, subset = Tage != "0")
```

## Add a cell cycle metadata

```{r}
###Add Cell Cycle
#From Seurat vignette at:
# https://satijalab.org/seurat/cell_cycle_vignette.html#assign-cell-cycle-scores
# Currently at least, the files are stored here:
# https://www.dropbox.com/s/3dby3bjsaf5arrw/cell_cycle_vignette_files.zip?dl=1
# `The file for the readtable line can be downloaded from the Seurat cell cycle vignette.
exp.mat <- read.table(file = "nestorawa_forcellcycle_expressionMatrix.txt",
    header = TRUE, 
    as.is = TRUE,
    row.names = 1)
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
#Scoring the cells:
Tcells <- CellCycleScoring(Tcells, s.features = s.genes, g2m.features = g2m.genes, 
    set.ident = TRUE)
# Store as metadata.
Tcells@meta.data$CellCycle <- Idents(Tcells)
Idents(Tcells) <- "Lane"
```

## Visualize after filtering

```{r}
multi_dittoPlot(Tcells,
    c("nFeature_RNA", "nCount_RNA","percent.mito"),
    group.by = "orig.ident",
    plots = "vlnplot")
# Better examined with:
dittoScatterPlot(Tcells,
                 "nFeature_RNA", "nCount_RNA", color.var = "percent.mito",
                 max = 20) +
    geom_hline(yintercept = 1500) +
    geom_vline(xintercept = 750)
demux.SNP.summary(Tcells,
    plots = c("jitter","vlnplot","boxplot"),
    boxplot.color = "white",
    boxplot.fill = F,
    add.line = 50)
hold <- Tcells$Lane
Tcells@meta.data$Lane <- factor(
  Tcells$Lane,
  levels = c("CD4","CD4-8","CD8"),
  labels = c("Lane1", "Lane2", "Lane3"))
demux.calls.summary(Tcells, xlab = NULL) +
    theme(axis.text.x = element_text(size = 9)) +
    theme(axis.text.y = element_text(size = 9)) +
    theme(axis.title.y = element_text(size = 10)) +
    theme(plot.title = element_text(size = 10))
Tcells@meta.data$Lane <- hold
table(Tcells$Sample,Tcells$Lane,Tcells$Tcelltype)
```

# 5. Data Pre-processing

## Log Transform / ("Normalize") the data
```{r}
Tcells <- NormalizeData(
  object = Tcells,
  normalization.method = "LogNormalize",
  scale.factor = 10000,
  verbose = F)
```

## Pick variable genes that will be used for PCA & tSNE
```{r}
Tcells <- FindVariableFeatures(object = Tcells, verbose = F, nfeatures = 2000)
```

## Make scaled data for the variable genes, regressing out certain variables

```{r}
#Scale without regressing out any variables
Tcells <- ScaleData(
  object = Tcells,
  vars.to.regress = c("CellCycle", "percent.mito", "nCount_RNA"),
  verbose = F)
```

# 6. Run PCA

```{r}
#Run PCA
Tcells <- RunPCA(
  object = Tcells,
  verbose = T,
  npcs = 50)
```

## Use PCheatmaps, PCElbowPlot, and JackStrawPlot to pick which PCs to carry forward into tSNE and clustering

```{r}
# PCHeatmap(
#     object = Tcells,
#     dims = 1:18, 
#     cells = 2000, 
#     balanced = TRUE
# )
# ElbowPlot(object = Tcells)
# #WARNING: JackStraw takes a long time to run.
# Tcells <- JackStraw(object = Tcells, num.replicate = 100, dims = 20, verbose = TRUE)
# Tcells <- ScoreJackStraw(Tcells, dims = 1:20)
# JackStrawPlot(object = Tcells, dims = 1:20)
```

The large change in p value magnitude from 9 to 10, and 9 pcs being around an inflection point in the elbow plot went into our picking PCs1-9 for tSNE, umap, and clustering.

# 7. Run tSNE
```{r tSNE}
Tcells <- RunTSNE(
  object = Tcells,
  reduction.use = "pca",
  dims = 1:9,
  seed.use = 1)

dittoDimPlot(Tcells, "Tage", size = 2, reduction.use = "tsne")
```

# 8. Run clustering
```{r clustering, echo=T}
Tcells <- FindNeighbors(Tcells,
                        reduction = "pca",
                        k.param = 20,
                        dims = 1:9)
Tcells <- FindClusters(Tcells,
                       modularity.fxn = 1,
                       algorithm = 1,
                       resolution = 0.1)
```

# 9. Run UMAP
```{r umap, fig.height = 5, fig.width = 6}
Tcells <- RunUMAP(
  object = Tcells,
  reduction = "pca",
  dims = 1:9,
  umap.method = "umap-learn",
  seed.use = 1)
```

```{r}
#Colored with CD4s lighter and CD8s darker 
dittoDimPlot(Tcells, "Tage", size = 1, reduction.use = "umap",
          colors = c(1:3,9:11), main = "T cells Lineage and Stage",
          rename.var.groups = c("Adult-CD4", "Fetal-CD4", "UCB-CD4",
                            "Adult-CD8", "Fetal-CD8", "UCB-CD8"))

plots <- list(
    dittoDimPlot(Tcells,
        "Tage", size = 0.5, reduction.use = "umap", cells.use = Tcells$Tage=="4-adult",
        colors = 1, legend.show = FALSE, ylab = NULL, xlab = NULL, main = NULL,
        show.axes.numbers = FALSE),
    dittoDimPlot(Tcells,
        "Tage", size = 0.5, reduction.use = "umap", cells.use = Tcells$Tage=="4-cord",
        colors = 2, legend.show = FALSE, ylab = NULL, xlab = NULL, main = NULL,
        show.axes.numbers = FALSE),
    dittoDimPlot(Tcells,
        "Tage", size = 0.5, reduction.use = "umap", cells.use = Tcells$Tage=="4-fetal",
        colors = 3, legend.show = FALSE, ylab = NULL, xlab = NULL, main = NULL,
        show.axes.numbers = FALSE),
    dittoDimPlot(Tcells,
        "Tage", size = 0.5, reduction.use = "umap", cells.use = Tcells$Tage=="8-adult",
        colors = 9, legend.show = FALSE, ylab = NULL, xlab = NULL, main = NULL,
        show.axes.numbers = FALSE),
    dittoDimPlot(Tcells,
        "Tage", size = 0.5, reduction.use = "umap", cells.use = Tcells$Tage=="8-cord",
        colors = 10, legend.show = FALSE, ylab = NULL, xlab = NULL, main = NULL,
        show.axes.numbers = FALSE),
    dittoDimPlot(Tcells,
        "Tage", size = 0.5, reduction.use = "umap", cells.use = Tcells$Tage=="8-fetal",
        colors = 11, legend.show = FALSE, ylab = NULL, xlab = NULL, main = NULL,
        show.axes.numbers = FALSE),
    dittoDimPlot(Tcells,
        "Tage", size = 1, reduction.use = "umap", legend.show = FALSE,
        color.panel = dittoColors()[c(1:3,9:11)], main = NULL),
    dittoSeq:::.grab_legend(dittoDimPlot(Tcells,
        "Tage", size = 1, reduction.use = "umap",
        color.panel = dittoColors()[c(1:3,9:11)],
        rename.var.groups = c("Adult-CD4", "UCB-CD4", "Fetal-CD4", 
                              "Adult-CD8", "UCB-CD8", "Fetal-CD8")))
)
pdf("Tcells-Figs/Tcell_umap_surround.pdf", w=6, h=6)
gridExtra::grid.arrange(grobs = plots,
    layout_matrix = matrix(c(
        7,7,7,4,
        7,7,7,5,
        7,7,7,6,
        1,2,3,8), ncol = 4))
dev.off()
```

#10 Quantify Clustering

```{r}
age.clust <- factor(Tcells$RNA_snn_res.0.1, labels = c("adult-cluster", "fetal-cluster", "UCB-cluster"))
Tcells@meta.data$age.cluster <- as.character(age.clust)
Idents(Tcells) <- "age.cluster"

pdf("Tcells-Figs/Clustering-umap_small.pdf", width = 3, height = 1.5)
dittoDimPlot(Tcells, "ident", size=0.3, do.label = FALSE, main = NULL, colors = c(1,3,2),
    xlab = NULL, ylab = NULL, show.axes.numbers = FALSE, reduction.use = "umap",
    rename.var.groups = c("adult-cluster", "fetal-cluster", "UCB-cluster"))
dev.off()

pdf("Tcells-Figs/Clustering-4s.pdf", width = 4, height = 1.7)
dittoBarPlot(Tcells, "ident", group.by = "Sample",
          x.reorder = c(6:10,11:15,1:5),
          main = NULL,
          cells.use = Tcells$Tcelltype=="CD4",
          ylab = "Fraction of CD4\nin each cluster",
          legend.show = FALSE, legend.title = "Clusters",
          x.labels = c(paste0("F",1:5),paste0("U",1:5),paste0("A",1:5)),
          x.labels.rotate = T,
          xlab = NULL)
dev.off()

pdf("Tcells-Figs/Clustering-8s.pdf", width = 4, height = 1.7)
dittoBarPlot(Tcells, "ident", group.by = "Sample",
          x.reorder = c(6:8,9:13,1:5),
          main = NULL,
          cells.use = Tcells$Tcelltype=="CD8",
          ylab = "Fraction of CD8\nin each cluster",
          legend.show = FALSE, legend.title = "Clusters",
          x.labels = c(paste0("F",1:5),paste0("U",1:5),paste0("A",1:5))[c(1,3,5:15)],
          x.labels.rotate = T,
          xlab = NULL)
dev.off()

# Percentage of fetal cells in fetal-cluster
sum(meta("ident",Tcells)=="fetal-cluster" & Tcells$age=="fetal") / sum(Tcells$age=="fetal")

# Percentage of UCB cells in UCB-cluster
sum(meta("ident",Tcells)=="UCB-cluster" & Tcells$age=="cord") / sum(Tcells$age=="cord")

# Percentage of adult cells in adult-cluster
sum(meta("ident",Tcells)=="adult-cluster" & Tcells$age=="adult") / sum(Tcells$age=="adult")

# Total percentage matching their age
(sum(meta("ident",Tcells)=="fetal-cluster" & Tcells$age=="fetal") +
    sum(meta("ident",Tcells)=="UCB-cluster" & Tcells$age=="cord") +
    sum(meta("ident",Tcells)=="adult-cluster" & Tcells$age=="adult")) /
    length(colnames(Tcells))

# Total percent UCB samples' cells in UCB cluster, CD4s
x <- table(
  Tcells$Sample[Tcells$age=="cord" & Tcells$Tcelltype=="CD4"],
  Tcells$age.cluster[Tcells$age=="cord" & Tcells$Tcelltype=="CD4"])
x[,2]/rowSums(x)

# Total percent UCB samples' cells in UCB cluster, CD8s
y <- table(
  Tcells$Sample[Tcells$age=="cord" & Tcells$Tcelltype=="CD8"],
  Tcells$age.cluster[Tcells$age=="cord" & Tcells$Tcelltype=="CD8"])
y[,2]/rowSums(y)

# Mean of number of cells per UCB sample, per T cell lineage, clustering as UCB 
mean(c(x[,2]/rowSums(x), y[,2]/rowSums(y)))
```

#11 Output differential expression

```{r}
Idents(Tcells) <- "age"
FvA <- FindMarkers(Tcells,
                       ident.1 = "fetal",
                       ident.2 = "adult",
                       test.use = "MAST")
FvA_padjFC <- FvA[abs(FvA$avg_logFC)>=0.585 &
                            FvA$p_val_adj<0.05 &
                            !(is.na(FvA$p_val_adj)),]
FvC <- FindMarkers(Tcells,
                       ident.1 = "fetal",
                       ident.2 = "cord",
                       test.use = "MAST")
FvC_padjFC <- FvC[abs(FvC$avg_logFC)>=0.585 &
                            FvC$p_val_adj<0.05 &
                            !(is.na(FvC$p_val_adj)),]
CvA <- FindMarkers(Tcells,
                       ident.1 = "cord",
                       ident.2 = "adult",
                       test.use = "MAST")
CvA_padjFC <- CvA[abs(CvA$avg_logFC)>=0.585 &
                           CvA$p_val_adj<0.05 &
                            !(is.na(CvA$p_val_adj)),]
write.csv(FvA_padjFC, row.names = TRUE, quote = FALSE, file = "Tcells_DiffExp_scFetalvsAdult.csv")
write.csv(FvC_padjFC, row.names = TRUE, quote = FALSE, file = "Tcells_DiffExp_scFetalvUCB.csv")
write.csv(CvA_padjFC, row.names = TRUE, quote = FALSE, file = "Tcells_DiffExp_scUCBvAdult.csv")
```

# 12 Developmental Stage Scoring

Steps:

1. Pick out a 10% of fetal and adult cells training set
2. Calculate the FvA markers for that set
3. Run correlation and random-forest feseR to narrow down the genelist.
4. Generate RFmodels based on feseR-restricted genesets
5. Check accuracy in fetal vs adult cells that were not in the training set
6. Score UCB (refered to as "cord" within the object)

```{r}
set.seed(1909)

### 0. Set ident to age.
Idents(Tcells) <- "age"

### 1. Pick 10% training group
inTraining <- createDataPartition(Idents(Tcells), p=0.1, list = FALSE)
#Trim to adult and fetal indices
inTraining <- inTraining[Idents(Tcells)[inTraining]%in%c("fetal","adult")]
sum(Tcells$age[inTraining]=="adult")/sum(Tcells$age[inTraining]%in%c("fetal", "adult"))
sum(meta("ident",Tcells)=="adult")/sum(meta("ident",Tcells)%in%c("fetal", "adult"))

### 2. Obtain markers
# Set Idents to 0 for non-training cells
age.inTrain <- Tcells$age
age.inTrain[-inTraining] <- 0
# age.inTrain[inTraining]
Idents(Tcells) <- age.inTrain
# Obtain markers
FvA <- FindMarkers(Tcells,
                       ident.1 = "fetal",
                       ident.2 = "adult",
                       test.use = "MAST")
FvA_padjFC <- FvA[abs(FvA$avg_logFC)>=0.585 &
                            FvA$p_val_adj<0.05 &
                            !(is.na(FvA$p_val_adj)),]
markers <- rownames(FvA_padjFC)

### 3. Run FESER to recursively eliminate features
#Grab data
training <- as.matrix(t(GetAssayData(Tcells)[markers,inTraining]))
# Creating the outcome score values
Train_val <- array(1, length(inTraining))
Train_val[Tcells$age[inTraining]=="fetal"] <- 0
#Run feseR correlation step
training.trim <- filter.corr(scale(training), Train_val, mincorr = 0.3)
#Run feseR random forest step
feser <- rfeRF(
  features = training.trim,
  class = Train_val,
  number.cv = 10,
  group.sizes = seq_len(ncol(training.trim)),
  metric = "ROC",
  verbose = FALSE)
#Pick optimal feser variables
(markers.feser <- feser$optVariables)
```

There are more than 20 features still.

```{r}
#Explore feseR results
feser$results
```

But, we can trim down to 17 and still keep ROC, sensitivity, and specificity above 0.99.

```{r}
### Extract the optVariables for using 17 total...
# Count how many of the repeated runs each feature ended up in.
vars17 <- unique(feser$variables$var[feser$variables$Variables==17])
(vars17.counts <- sapply(vars17, function(X) length(grep(X, feser$variables$var[feser$variables$Variables==17]))))
```

```{r}
# Extract the ones that showed up the most
(vars.use <- names(head(vars17.counts[order(vars17.counts, decreasing = TRUE)], 17)))
```

```{r}
markers.feser <- vars.use

### 4. Generate final RF model
#Grab data
training <- as.matrix(t(GetAssayData(Tcells)[markers.feser,inTraining]))
# Creating the outcome score values
Train_val <- array(1, length(inTraining))
Train_val[Tcells$age[inTraining]=="fetal"] <- 0
# Make the model
rf_mod <- train(Train_val ~ .,
              set.seed(998),
              data= cbind(training,Train_val),
              method = "ranger",
              metric = "MAE",
              trControl = trainControl(method = "cv",
                                       number = 3,
                                       repeats = 3),
              tuneGrid = expand.grid(mtry = round(length(markers.feser)*.75,0),
                                     splitrule = c("extratrees"),
                                     min.node.size = 1)
              )

### 5. Score all
Tcells@meta.data$RFScore <- as.double(predict(rf_mod,t(GetAssayData(Tcells)[markers.feser,])))
Tcells@meta.data$inTraining <- FALSE
Tcells@meta.data$inTraining[inTraining] <- TRUE

### 6. Score FvA accuracy
# Set ident back to full age.
Idents(Tcells) <- "age"
roc_obj <- roc(response = as.numeric(meta("ident",Tcells)[!(Tcells$inTraining) &
                                                   meta("ident",Tcells)%in%c("fetal", "adult")]=="adult"),
               predictor = Tcells$RFScore[!(Tcells$inTraining) &
                                               meta("ident",Tcells)%in%c("fetal", "adult")],
               plot = T)
auc(roc_obj)
```

```{r}
dittoPlot(Tcells, "RFScore", cells.use = Tcells$inTraining,
       group.by = "Sample", color.by = "age",
       plots = c("jitter","vlnplot"),
       boxplot.color = "white", boxplot.fill = F,
       vlnplot.lineweight = 0.3, vlnplot.width = 3,
       sub = "in Training", colors = c(1,3))

### 7. Check the look for all T cells.
dittoPlot(Tcells, "RFScore", cells.use = !(Tcells$inTraining),
       group.by = "Sample", color.by = "age",
       plots = c("jitter","vlnplot"),
       boxplot.color = "white", boxplot.fill = F,
       vlnplot.lineweight = 0.3, vlnplot.width = 5,
       sub = "NOT in training")
dittoDimPlot(Tcells, "RFScore", cells.use = !(Tcells$inTraining),
             sub = "NOT in training", size = 2, reduction.use = "umap")
```


```{r}
pdf("Tcells-Figs/Tcell_train-umap.pdf", w=1.38, h=1.3)
dittoDimPlot(Tcells, "age", colors = c(1,3), cells.use = Tcells$inTraining, main = NULL,
             xlab = NULL, ylab = NULL, legend.show = FALSE, size = 0.3, reduction.use = "umap",
             show.axes.numbers = FALSE)
dev.off()
pdf("Tcells-Figs/Tcell_extend-umap.pdf", w=3.1, h=1.7)
dittoDimPlot(Tcells, "RFScore", colors = c(1,3), main = NULL,
             xlab = NULL, ylab = NULL, legend.show = TRUE, size = 0.5,
             legend.breaks = c(0,0.5,1), legend.breaks.labels = c("0, fetal-like", "0.5", "1, adult-like"),
             reduction.use = "umap", show.axes.numbers = FALSE)
dev.off()
```

```{r}
pdf("Tcells-Figs/RFScore-4s.pdf", width = 6, height = 3)
dittoPlot(Tcells, "RFScore", group.by = "Sample", color.by = "age",
       cells.use = Tcells$Tcelltype=="CD4",
       plots = c("jitter","vlnplot"),
       boxplot.color = "white", boxplot.fill = F, boxplot.width = 0.4,
       jitter.size = 0.3,
       vlnplot.lineweight = 0.3, vlnplot.width = 7,
       x.labels = c(paste0("Fetal-",1:5),paste0("UCB-",1:5),paste0("Adult-",1:5)),
       x.reorder = c(6:15,1:5),
       y.breaks = seq(0,1,.5),
       legend.show = FALSE,
       xlab = NULL,
       ylab = "Developmental Stage Score",
       main = "Developmental Stage Score - Naive CD4 T cells"
       )
dev.off()
pdf("Tcells-Figs/RFScore-8s.pdf", width = 6, height = 3)
dittoPlot(Tcells, "RFScore", group.by = "Sample", color.by = "age",
       cells.use = Tcells$Tcelltype=="CD8",
       plots = c("jitter","vlnplot"),
       boxplot.color = "white", boxplot.fill = F, boxplot.width = 0.4,
       jitter.size = 0.3,
       vlnplot.lineweight = 0.3, vlnplot.width = 7,
       x.labels = c(paste0("Fetal-",1:5),paste0("UCB-",1:5),paste0("Adult-",1:5))[c(1,3,5:15)],
       x.reorder = c(6:13,1:5),
       y.breaks = seq(0,1,.5),
       legend.show = FALSE,
       xlab = NULL,
       ylab = "Developmental Stage Score",
       main = "Developmental Stage Score - Naive CD8 T cells"
       )
dev.off()
```

## Scores Summary

```{r}
# Scores less than 0.1
  # fetal
sum((Tcells$RFScore<0.1 & Tcells$age=="fetal")) / sum(Tcells$age=="fetal")
  # cord
sum((Tcells$RFScore<0.1 & Tcells$age=="cord")) / sum(Tcells$age=="cord")
  # adult
sum((Tcells$RFScore<0.1 & Tcells$age=="adult")) / sum(Tcells$age=="adult")

# Scores between 0.1 and 0.9
  # fetal
sum((Tcells$RFScore>=0.1 & Tcells$RFScore<=0.9 & Tcells$age=="fetal")) / sum(Tcells$age=="fetal")
  # cord
sum((Tcells$RFScore>=0.1 & Tcells$RFScore<=0.9 & Tcells$age=="cord")) / sum(Tcells$age=="cord")
  # adult
sum((Tcells$RFScore>=0.1 & Tcells$RFScore<=0.9 & Tcells$age=="adult")) / sum(Tcells$age=="adult")

# Scores greater than 0.9
  # fetal
sum((Tcells$RFScore>0.9 & Tcells$age=="fetal")) / sum(Tcells$age=="fetal")
  # cord
sum((Tcells$RFScore>0.9 & Tcells$age=="cord")) / sum(Tcells$age=="cord")
  # adult
sum((Tcells$RFScore>0.9 & Tcells$age=="adult")) / sum(Tcells$age=="adult")

# mean & stdev
  # fetal
mean(Tcells$RFScore[Tcells$age=="fetal"])
sd(Tcells$RFScore[Tcells$age=="fetal"])
  # cord
mean(Tcells$RFScore[Tcells$age=="cord"])
sd(Tcells$RFScore[Tcells$age=="cord"])
  # adult
mean(Tcells$RFScore[Tcells$age=="adult"])
sd(Tcells$RFScore[Tcells$age=="adult"])
```

## RF Model Markers expression comparison

```{r}
# Reorder to put UCB~fetal-like genes first and UCB~adult-like genes last
markers.feser <- markers.feser[c(1,9,12,15,2:4,6:8,10,11,14,17,13,16,5)]
pdf("Tcells-Figs/RFmarkers_Tcells.pdf", w = 7.5, h = 6.5)
multi_dittoPlot(Tcells, markers.feser, group.by = "age", ncol = 5, nrow = 4,
    x.reorder = 3:1, x.labels = c("F","U","A"), x.labels.rotate = FALSE,
    vlnplot.lineweight = 0.3, jitter.size = 0.2)
dev.off()
```

# 13 SAVE
```{r}
saveRDS(Tcells, file = "Tcells.rds")
```